A-maze of Natural Stories: Texts are comprehensible using the Maze task

### 2022-10-26 Papers/Paper/manuscript.Rmd
first pass at edits for R&R

However, there is increasing interest in collecting data during comprehension of more naturalistic materials such as stories and news articles [@demberg-keller:2008; @lukeLimitsLexicalPrediction2016; @futrellNaturalStoriesCorpus2020], which offer potentially improved ecological validity and larger scale data in comparison with repeated presentation of isolated sentences out of context. These more naturalistic materials require maintaining and integrating discourse dependencies and other types of information over longer stretches of time and linguistic material. Previous work leaves unclear whether the Maze task would be feasible for this purpose: the increased task demands might interfere with the demands presented by these more naturalistic materials, and vice versa. In this paper we report a new modification of the Maze task and show that it makes reading of extended, naturalistic texts feasible. We also analyze the resulting reaction time profiles and show that they provide strong signal regarding the probabilistic relationship between a word and the context in which it appears, and that the systematic linear relationship between word surprisal and response time observed in other reading paradigms [@smithEffectWordPredictability2013] also arises in the Maze task.
<!-- Behavioral measures of word-by-word reading time provide experimental evidence to test theories of language processing. Maze is a potentially promising method, and it's feasibility has recently increased due to the introduction of A-Maze [@boyceMazeMadeEasy2020]. -->
In the remainder of the Introduction, we lay out the role of RT-based methods in theory testing, describe a few common methods, and review some key influences on reading time. We then proceed to present our modified "error-correction Maze" paradigm, our experiment, and the results of our analyses of the resulting data.

@@

For instance, competing theories about why certain types of object-extracted relative clauses, like *the lawyer that the banker irritated*, are harder to understand than the corresponding subject-extracted relative clauses, like *the lawyer that irritated the banker*, make different predictions about which words are the loci of the overall difficulty and slower RTs associated with object relatives [@grodnerConsequencesSerialNature2005; @staubEyeMovementsProcessing2010; @traxlerProcessingSubjectObject2002]. RT measures can potentially also inform theories about the time course of processing (i.e. which steps are parallel versus serial, @bartekSearchOnlineLocality2011) or the functional form of relationships between word characteristics and processing time [@smithEffectWordPredictability2013].

Some of these theories rely on being able to attribute processing slowdowns to a particular word. Determining that object relatives are overall slower that subject relatives is easy. Even an imprecise RT measure will determine that the same set of words in a different order took longer to read at a sentence level. However, many language processing theories make specific (and contrasting) theories about which words in a sentence are harder to process. To adjudicate among these theories, we want methods that are *well-localized*, so it is easy to determine which word is responsible for an observed RT slow-down. Ideally, longer RT on a word would be an indication of that word's increased difficulty, and not the lingering signal of a prior word's increased difficulty. When the signal isn't localized, advanced analysis techniques may be required to disentangle the slow-downs [@shainDeconvolutionalTimeSeries2018].

@@

In eye-tracking, participants read a text on a screen naturally, while their saccadic eye movements are recorded on a computer-connected camera that is calibrated so that the researcher can reconstruct with high precision where the participant's gaze falls on the screen at all times [@raynerEyeMovementsReading1998]. These eye movements can be used to reconstruct various position-specific reading time measures such as *gaze duration* (the total amount of time the eyes spend on a word the first time it is fixated, or zero if the eye skipped the word the first time it was approached from the left) and *total viewing time* (the total amount of time that the word is fixated). Eye tracking data collected with state-of-the-art high-precision recording equipment offers relatively good signal-to-noise ratio, but the difficulty presented by a word can still *spill over* into reading measures on subsequent words, a dynamic that can make it hard to isolate the source of an effect of potential theoretical interest [@raynerEffectsFrequencyPredictability2004; @levyEyeMovementEvidence2009; @frazierMakingCorrectingErrors1982]. Additionally, the equipment is expensive and data collection is laborious and must occur in-lab.

@@

@@

The overall predictability of a word in a context can be estimated using language models that are trained on large corpora of language to predict what word comes next in a sentence. A variety of pre-trained models exist, with varied internal architectures and training methods, but all of them generate measures of predictability. Predictability is often measured in bits of surprisal, which is the negative log probability of a word (@hale:2001; @levy:2008; 1 bit of surprisal means a word is expected to occur half the time, 2 bits is 1/4 of the time, etc.).

@@

@@

We used the Natural Stories corpus [@futrellNaturalStoriesCorpus2020], which consists of 10 passages of roughly 1000 words each which are designed to read fluently to native speakers. At the same time, the passages contain copious punctuation, quoted speech, proper nouns, and low frequency grammatical constructions. The corpus is accompanied by binary-choice comprehension questions, 6 per story, which we used to assess comprehension.

@@

## Task: Error-correction Maze

@@

## Procedure
Participants first gave their informed consent and saw task instructions. Then they read a short practice story in the Maze paradigm and answered 2 binary-choice practice comprehension questions, before reading one main story in the A-maze task. After the story, they answered 6 comprehension questions, commented on their experience, answered optional demographic questions, were debriefed, and were given a code to enter for payment. The experiment was implemented in Ibex (https://github.com/addrummond/ibex).

@@

## SPR--Maze correlation
We compared the correlations between the Maze and SPR RTs to within-Maze and within-SPR correlations. For Maze, within each story, we randomly split subjects into two halves. Within each half, we calculated a per-word average RT for each word and then a per-sentence average RT across word averages. We calculated a within-Maze correlation between these two halves.

For this comparison, we downsampled the SPR data choosing a number of participants equal to the number we have for Maze to avoid differences due to dataset size. We then used the same procedure to get a within-SPR correlation. For between Maze-SPR correlation, we took the average correlation across each of the 4 pairs of Maze half and SPR half.

@@

Our analytic questions required multiple modeling approaches. To look at the functional form of the relationship between surprisal and RT data, we fit Generalized Additive Models (GAMs) to allow for non-linear relationships. GAM model summaries can be harder to interpret than those for linear models, so to measure effect sizes and assess spillover, we used linear mixed models. Finally, in order to determine which language model best predicts the RT data, we fit additional linear models with predictors from multiple language models to look at their relative contributions. All these models used surprisal, frequency, and length as predictors for RT. We considered these predictors from both the current and past word to account for the possibility of spillover effects in A-maze. For SPR comparisons, we included predictors from the current and past three words to account for known spillover effects. We conducted data processing and analyses using `r a$r`.

@@

We excluded the first word of every sentence because it had an x-x-x distractor, leaving `r nrow(not_first)` words. We excluded words for which we didn't have surprisal or frequency information, leaving `r nrow(not_na)` words. We additionally excluded words that any model treated as being composed of multiple tokens (primarily words with punctuation),  leaving `r nrow(not_gpt)` words[^1]. We excluded outlier RTs that were <100 or >5000 ms (<100 is likely a recording error, >5000 is likely the participant getting distracted). We exclude RTs from words where mistakes occurred or which occurred after a mistake in the same sentence. We only analyzed words where we had values for all predictors, which meant that if the previous word was unknown to a model, the word was excluded because of missing values for a lagged predictor.

@@

## How do A-maze and SPR compare in power and reliability?

@@

The Maze linear model effects are shown in Figure \@ref(fig:coeffs-maze) (See also Appendix B for a table of effects). Across all models, there were consistent large effects of length and surprisal at the current word, but minimal effects of frequency. The lack of frequency effects is unexpected, but consistent with @shainLargescaleStudyEffects2019. There was a small interaction between surprisal and length at the current word.

@@

Second, we took advantage of the pre-existing SPR corpus on Natural Stories to compare the RT profiles between Maze and SPR.  Maze and SPR pick up on similar features in words, as shown by the high correlations between Maze and SPR RTs on the sentence level.

@@

Fourth, we compared the spillover profiles between Maze and SPR. For Maze, we found large effects of the current word's surprisal and length, which dwarfed any spillover effects from previous word predictors. In contrast, for SPR, we found effects of roughly equal sizes from the current and previous words. ^[Furthermore, the typical spillover profile for SPR data may be worse than suggested by the Natural Stories corpus SPR data: for example, @smithEffectWordPredictability2013 found that most of a word's surprisal effect showed up only one to two words downstream.] Overall, Maze is a slower task than SPR, but it also has much larger effects of length and surprisal, perhaps due to requiring more focus, and thus generating less noisy data.

@@

Overall, A-maze has excellent localization, although some  models showed small but statistically reliable effects of the past word. On the whole, our results support the idea that Maze forces language processing to be close to word-by-word, and thus the Maze task can be used under the assumption that the RT of a word primarily reflects its own properties and not those of earlier words.

@@

Compared to traditional Maze, error-correction Maze reduces perverse (from the experimenter's point of view) incentives to complete the task as quickly as possible. However, even with error-correction Maze, clicking randomly is still likely faster than doing the task. In discussing this work, we received the suggestion that one way to further disincentivize random clicking would be to add a pause when a participant makes a mistake, forcing them to wait some short period of time (ex 500ms) before correcting their mistake. This delay would make randomly hitting buttons slower than doing the task as intended, and we have made delaying after wrong presses an option in the error-correction Maze implementation at link obscured for anonymity <!--https://github.com/vboyce/Ibex-with-Maze-->.

>>>>

There is increasing interest in collecting data during comprehension of more naturalistic materials such as stories and news articles [@demberg-keller:2008; @lukeLimitsLexicalPrediction2016; @futrellNaturalStoriesCorpus2020], which offer potentially improved ecological validity and larger scale data in comparison with repeated presentation of isolated sentences out of context. These more naturalistic materials require maintaining and integrating discourse dependencies and other types of information over longer stretches of time and linguistic material. Previous work leaves unclear whether the Maze task would be feasible for this purpose: the increased task demands might interfere with the demands presented by these more naturalistic materials, and vice versa. In this paper we report a new modification of the Maze task and show that it makes reading of extended, naturalistic texts feasible. We also analyze the resulting reaction time profiles and show that they provide strong signal regarding the probabilistic relationship between a word and the context in which it appears, and that the systematic linear relationship between word surprisal and response time observed in other reading paradigms [@smithEffectWordPredictability2013] also arises in the Maze task.
<!-- Behavioral measures of word-by-word reading time provide experimental evidence to test theories of language processing. Maze is a potentially promising method, and it's feasibility has recently increased due to the introduction of A-Maze [@boyceMazeMadeEasy2020]. -->
In the remainder of the Introduction, we lay out the role of RT-based methods in theory testing, describe a few common methods, and review some key influences on reading time. We then proceed to present our modified "error-correction Maze" paradigm, our experiment, and the results of our analyses of the resulting data.

@@

For instance, competing theories about why certain types of object-extracted relative clauses, like *the lawyer that the banker irritated*, are harder to understand than the corresponding subject-extracted relative clauses, like *the lawyer that irritated the banker*, make different predictions about which words are the loci of the overall difficulty and slower RTs associated with object relatives  [@traxlerProcessingSubjectObject2002]. Dependency-locality theory [@grodnerConsequencesSerialNature2005] predicts that the locus of difficulty on object relatives is on the verb of the embedded clause (*irritated* in *the lawyer that the banker irritated*). In contrast, surprisal theory predicts the locus of difficulty will be on the article *the* at the start of the embedded clause [@staubEyeMovementsProcessing2010]. RT measures can potentially also inform theories about the time course of processing (i.e. which steps are parallel versus serial, @bartekSearchOnlineLocality2011) or the functional form of relationships between word characteristics and processing time [@smithEffectWordPredictability2013].

Some of these theories rely on being able to attribute processing slowdowns to a particular word. Determining that object relatives are overall slower that subject relatives is easy. Even an imprecise RT measure will determine that the same set of words in a different order took longer to read at a sentence level. However, many language processing theories make specific (and contrasting) predictions about which words in a sentence are harder to process. To adjudicate among these theories, we want methods that are *well-localized*, so it is easy to determine which word is responsible for an observed RT slow-down. Ideally, a longer RT on a word would be an indication of that word's increased difficulty, and not the lingering signal of a prior word's increased difficulty. When the signal isn't localized, advanced analysis techniques may be required to disentangle the slow-downs [@shainDeconvolutionalTimeSeries2018].

@@

In eye-tracking, participants read a text on a screen naturally, while their saccadic eye movements are recorded on a computer-connected camera that is calibrated so that the researcher can reconstruct with high precision where the participant's gaze falls on the screen at all times [@raynerEyeMovementsReading1998]. These eye movements can be used to reconstruct various position-specific reading time measures such as *gaze duration* (the total amount of time the eyes spend on a word the first time it is fixated before saccading to a later word) and *total viewing time* (the total amount of time that the word is fixated). If the eyes skipped the word the first time it was approached to the left, the trial is generally excluded.  Eye tracking data collected with state-of-the-art high-precision recording equipment offers relatively good signal-to-noise ratio, but the difficulty presented by a word can still *spill over* into reading measures on subsequent words, a dynamic that can make it hard to isolate the source of an effect of potential theoretical interest [@raynerEffectsFrequencyPredictability2004; @levyEyeMovementEvidence2009; @frazierMakingCorrectingErrors1982]. Short words such as articles and pronouns are often not fixated directly which makes it harder to study the processing of these words with eye-tracking. Additionally, the equipment is expensive and data collection is laborious and must occur in-lab.

@@

The Maze task has much less face-validity than eye-tracking or even SPR; repeated forced-choice selections does not look very similar to normal reading. Despite this, @forsterMazeTaskMeasuring2009 report that "At a phenomenological level, participants typically report that they feel as if they are reading the sentence relatively
naturally and that the correct alternative seems to “leap out” at them, so that they do not have to inspect the incorrect
alternative very carefully, if at all." This suggests that the Maze task is not about conscious comparison between two potential continuations and instead relies on the same language processing facilities shared by other methods. Maze is not trying to study the process of normal reading, but instead to get underlying language processing, using an artificial method to slow down and localize language processing.

@@

The overall predictability of a word in a context can be estimated using language models that are trained on large corpora of language to predict what word comes next in a sentence. A variety of pre-trained models exist, with varied internal architectures and training methods, but all of them generate measures of predictability. Predictability is often measured in bits of surprisal, which is the negative log probability of a word [@hale:2001, @levy:2008]. 1 bit of surprisal means a word is expected to occur half the time, 2 bits is 1/4 of the time, etc.

@@

Frequency is often found to be a predictor of reading time in SPR and eye-tracking experiments. It is unclear why we find frequency effects on RT when surprisal effects are accounted for. In their discussion, @smith-levy:2008 posit that the role of frequency could diminish or disappear as measurements of surprisal improve (as they have over the past decade). Additionally, @smith-levy:2008 hypothesize that the effects of frequency (in addition to those of surprisal) may be the result of quick, heuristic responses to the words based on the more easily available unigram frequency before more fine-grained context-sensitive surprisal is available. An alternative is that frequency effects don't actually exist, but are an artifact of how RT data is analysed [@shainLargescaleStudyEffects2019].

@@

We used the Natural Stories corpus [@futrellNaturalStoriesCorpus2020], a set of 10 passages designed to read fluently to native speakers. Each passage is roughly 1000 words long. The passages contain copious punctuation, quoted speech, proper nouns, and low frequency grammatical constructions. The corpus is accompanied by binary-choice comprehension questions, 6 per story, which we used to assess comprehension.

@@

# Error-correct Maze: Adapting the Maze paradigm for long passages

@@

## Procedure
Participants first gave their informed consent and saw task instructions. Then they read a short practice story in the Maze paradigm and answered 2 binary-choice practice comprehension questions, before reading one main story in the error-correction A-maze task. After the story, they answered 6 comprehension questions, commented on their experience, answered optional demographic questions, were debriefed, and were given a code to enter for payment. The experiment was implemented in Ibex (https://github.com/addrummond/ibex).

@@

## SPR--Maze correlation
We compared the correlations between the Maze and SPR RTs to within-Maze and within-SPR correlations. For Maze, within each story, we split the data in half, randomly assigning subjects into two equal groups. Within each half, we calculated a per-word average RT for each word and then a per-sentence average RT across word averages. We calculated a within-Maze correlation between these two halves.

For this comparison, we downsampled the SPR data choosing a number of participants equal to the number we have for Maze to avoid differences due to dataset size. We then used the same split-half procedure to get a within-SPR correlation. For between Maze-SPR correlation, we took the average correlation across each of the 4 pairs of Maze half and SPR half.

@@

Our analytic questions required multiple modeling approaches. To look at the functional form of the relationship between surprisal and RT data, we fit Generalized Additive Models (GAMs) to allow for non-linear relationships [@wood:2006GAMs]. GAM model summaries can be harder to interpret than those for linear models, so to measure effect sizes and assess spillover, we used linear mixed models. Finally, in order to determine which language model best predicts the RT data, we fit additional linear models with predictors from multiple language models to look at their relative contributions. All these models used surprisal, frequency, and length as predictors for RT. We considered these predictors from both the current and past word to account for the possibility of spillover effects in A-maze. For SPR comparisons, we included predictors from the current and past three words to account for known spillover effects. We conducted data processing and analyses using `r a$r`.

@@

In the Maze task, the first word of every sentence is paired with a nonce (x-x-x) distractor rather than a real word (as there is no context to use to distinguish between real words); due to this difference, we excluded the first word of every sentence, leaving `r nrow(not_first)` words. We excluded words for which we didn't have surprisal or frequency information, leaving `r nrow(not_na)` words. We additionally excluded words that any model treated as being composed of multiple tokens (primarily words with punctuation),  leaving `r nrow(not_gpt)` words[^1]. We excluded outlier RTs that were <100 or >5000 ms (<100 is likely a recording error, >5000 is likely the participant getting distracted). We exclude RTs from words where mistakes occurred or which occurred after a mistake in the same sentence. We only analyzed words where we had values for all predictors, which meant that if the previous word was unknown to a model, the word was excluded because of missing values for a lagged predictor.

@@

## How do A-maze and SPR compare in reliability?

@@

The Maze linear model effects are shown in Figure \@ref(fig:coeffs-maze) (See also Appendix B for a table of effects). Across all models, there were consistent large effects of length and surprisal at the current word, but minimal effects of frequency. This lack of frequency effects is inconsistent with the results usually reported for SPR and eye-tracking. There was a small interaction between surprisal and length at the current word.

@@

Second, we took advantage of the pre-existing SPR corpus on Natural Stories to compare the RT profiles between Maze and SPR.  Maze and SPR pick up on similar features in words, as shown by the high correlations between Maze and SPR RTs on the sentence level. The correlation within Maze is higher than Maze to SPR correlation or SPR-SPR correlations which is evidence that Maze is less noisy that SPR and may pick up on deeper linguistic features than SPR.

@@

Fourth, we compared the spillover profiles between Maze and SPR. For Maze, we found large effects of the current word's surprisal and length, which dwarfed any spillover effects from previous word predictors. In contrast, for SPR, we found effects of roughly equal sizes from the current and previous words. ^[Furthermore, the typical spillover profile for SPR data may be worse than suggested by the Natural Stories corpus SPR data: for example, @smithEffectWordPredictability2013 found that most of a word's surprisal effect showed up only one to two words downstream.] Overall, Maze is a slower task than SPR, but it also has much larger effects of length and surprisal, perhaps due to requiring more focus, and thus generating less noisy data. 
We do not find frequency effects on the Maze data, but we do on the SPR data. This could be explained if frequency effects are a first rough approximation of in-context predictability, before the fuller context-sensitive surprisal information is available. In this case, faster methods like eye-tracking and SPR would show frequency effects (in addition to surprisal), but slower methods like Maze would not as the additional demands slow down the response, allowing more contextual information to be used. While this is a difference between Maze and other incremental processing methods, we do not consider it a flaw, since there is not strong theoretical justification for the presence of frequency effects (in addition to surprisal effects) in language processing. These differences do point to the importance of understanding task demands of different incremental processing methods.

@@

Overall, A-maze has excellent localization, although some models showed small but statistically reliable effects of the past word. On the whole, our results support the idea that Maze forces language processing to be close to word-by-word, and thus the Maze task can be used under the assumption that the RT of a word primarily reflects its own properties and not those of earlier words. Correlations between Maze and SPR indicate that Maze is picking up on similar patterns, but with less noise, than SPR.

@@

Compared to traditional Maze,in error-correction Maze, participants' incentives to finish quickly are in less conflict with the experimenter's desire that participants do the task as intended. However, even with error-correction Maze, clicking randomly is still likely faster than doing the task. In discussing this work, we received the suggestion that one way to further disincentivize random clicking would be to add a pause when a participant makes a mistake, forcing them to wait some short period of time (ex 500ms) before correcting their mistake. This delay would make randomly hitting buttons slower than doing the task as intended, and we have made delaying after wrong presses an option in the error-correction Maze implementation at link obscured for anonymity <!--https://github.com/vboyce/Ibex-with-Maze-->.
